Design Document: Monte Carlo Maximum Likelihood for Generalized Linear Mixed Models

### 2014-11-24 NewDesignDoc/DesignDocGLMM.tex
Editing design doc

Then the Monte Carlo log likelihood approximation is
\begin{align}
l_{m}(\theta) &=\log \dfrac{1}{m} \sum_{k=1}^mf_\theta(y|u_k)  \dfrac{ f_\theta(u_k)   }{\tilde{f}(u_k)}\\
&= \log \dfrac{1}{m} \sum_{k=1}^m  \dfrac{ f_\theta(y,u_k)   }{\tilde{f}(u_k)}.
\end{align}
When $\tilde{f}$ depends on $\theta$, the gradient vector of the MCLA with respect to $\theta$ is
\begin{align}
\nabla l_m(\theta)&= \dfrac{\sum_{k=1}^m    \left( \nabla \log f_\theta(u_k,y) - \nabla \log \tilde{f} (u_k)  \right) \dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)} }{\sum_{k=1}^m \left( \dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)} \right) }, \label{eq:MCLAgradient}
\end{align}
and the Hessian matrix of the MCLA is
\begin{align}
&\nabla^2 l_m(\theta)= \dfrac{   \sum_{k=1}^m \left[ \nabla^2 \log f_\theta(y,u_k)  -
   \nabla^2 \log \tilde{f}(u_k)   \right]  \dfrac{ f_\theta(y,u_k)}{\tilde{f}(u_k)}  }{\sum_{k=1}^m  \dfrac{ f_\theta(y,u_k)   }{\tilde{f}(u_k)}}\\
&+ \dfrac{   \sum_{k=1}^m \left[ \nabla \log f_\theta(y,u_k)  -
   \nabla \log \tilde{f}(u_k) - \nabla l_m(\theta)   \right] \left[ \nabla \log f_\theta(y,u_k)  -
   \nabla \log \tilde{f}(u_k) -\nabla l_m(\theta)  \right]'  \dfrac{ f_\theta(y,u_k)   }{\tilde{f}(u_k)}   }{\sum_{k=1}^m  \dfrac{ f_\theta(y,u_k)   }{\tilde{f}(u_k)}}.
\end{align}
Details for these equations can be found in section \ref{sec:calcs}. Calculation for the gradient and hessian when $\tilde{f}$ is independent of $\theta$ can be found in section \ref{sec:calcsindep}.

@@

\section{Random effect generation and calculations}
This section is focused on a vector of random effects $u$ with length $q$. We assume that $u \sim N(0,D)$ where $D$ depends on $\nu$. Section \ref{sec:getEk} details the  relationship between  $\nu$ and $D$ explicit. Section \ref{sec:genRand} describes the process of generating the random effects. Section \ref{sec:distRandsec} explains how to evaluate the distribution of the random effects and its first two derivatives.

@@

If $q_t$ is the number of random effects associated with variance component $\nu_t$, then $q_t$ is also the number of nonzero entries in $E_t$ and $q=\sum_{t=1}^T q_t$ is the total number of random effects in the model.  
The \texttt{model.matrix} part of \texttt{glmm} has made  \texttt{Z} into a list with $T$ design matrices (one for each distinct variance component).  In order to create $E_t$, we need to go through and count the  number of columns ($q_t$) for each design matrix $t$ in the list.
 Then $q=\sum_{t=1}^T q_t$ is the total number of random effects in the model, and $D$ will have $q$ rows (and columns). Thus, each $E_t$ will have $q$ rows and columns as well.  Then we know that $E_1$ has $q_1$ ones on the diagonal, followed by zeros to fill the rest of the diagonal.  $E_2$ has $q_1$ zeros, then $q_2$ ones, then zeros for the rest of the diagonal.

\subsection{Generating random effects (genRand)}\label{sec:genRand}
The model fitting function \texttt{glmm}  calls the function that generates the random effects and  stores them for future calculations. Let $s$ be a vector of length $q$ that represents the random effects on the standard normal scale. That is, $u= D^{1/2}s$. Let $\sigma$ be a vector of length $T$ with components $\sqrt{\nu_t}$.  The  PQL step of \texttt{glmm} has resulted in PQL estimates for $\beta$, $\sigma$ and  $s$ denoted by $\beta^*$, $s^*$ and $\sigma^*$. Then let
\begin{align}
A^*=\sum_{t=1}^T E_t \sigma^*_t
\end{align}\
and
\begin{align}
 D^* =A^*A^*.
\end{align}
Also, we can ``unstandardize'' our random effects:
\begin{align}
u^*&=A^*s^*.
\end{align}

Let $p_1,p_2,p_3$ be proportions such that  $p_1+p_2+p_3=1$. Let $D$ be determined by the current iteration of \texttt{trust}. The importance sampling distribution $\tilde{f}$ is the following mixture distribution:
\begin{align}
 \tilde{f}(u) = p_1  f(u \, | \, 0, D)+p_2  f(u \, | \, u^*, D^*)+p_3  f(u \, | \, u^*, (Z'  c''(X \beta^*+Zu^*) Z +(D^*)^{-1}   )^{-1})
\end{align}

@@

The random effects are generated by shifting and scaling a sample from a  standard normal distribution.  Suppose we want to generate a sample from a general normal distribution $N(\mu,\Sigma)$.  Let $u_k$ be a vector with length $q$ and let it denote one vector of generated random effects.  For each k, draw a sample of size $q$ from the standard normal distribution. Call this sample $\breve{u}_k$. Then shift and scale it:
\begin{align}
u_k = \breve{u}_k \Sigma^{1/2} + \mu.
\end{align}
so that is a sample from $N(\mu,\Sigma)$.

One way to find $\Sigma^{1/2}$ is through eigendecomposition. Eigendecompositions take a little bit more time than Cholesky decompositions but are more stable. We can use eigendecomposition for $\Sigma$ to get orthogonal matrix $O$ (containing the eigenvectors) and diagonal matrix $\Lambda$ (with diagonal entries $\lambda$ being the eigenvalues of $\Sigma$).  Then
\begin{align}
\Sigma^{1/2}= O \Lambda^{1/2} O'.
\end{align}
Finding $\Lambda^{1/2}$ is as easy as taking the root of the diagonal entries.  We know that $\Sigma^{1/2}$ is correct because
\begin{align}
\Sigma^{1/2} \Sigma^{1/2} &= O \Lambda^{1/2} O'O \Lambda^{1/2} O'\\
&= O \Lambda^{1/2}\Lambda^{1/2} O'\\
&=O \Lambda O' \\
&= \Sigma.
\end{align}

@@

We assume the random effects are $N(\mu, \Sigma)$.  This is written with $\mu$ not necessarily $0$ so that the function can be used for $\tilde{f}(u)$ as well.  We can  write the distribution of the random effects as:
\begin{align}
\log f (u| \mu, \Sigma) = (-1/2) \log |\Sigma| - (1/2) (U-\mu)' \Sigma^{-1} (U-\mu)
\end{align}

For now, I  assume that  D is diagonal. Formulas for this are in section \ref{sec:Ddiag}.  I generate random effects both from a normal distribution with $\Sigma=D$ diagonal (details in \ref{sec:Ddiag}) and from a normal distribution with $\Sigma$ general (details in \ref{sec:Dgeneral}).

\subsubsection{$\Sigma=D$ Diagonal}\label{sec:Ddiag}
These equations will be used for both $\log f_\theta(u)$ and for pieces of $\tilde{f}(u)$.  This section provides the equations for finding the value  of a normal density with $D$ diagonal, the gradient vector, and the hessian matrix.

In the simplest case, $\Sigma=D$ is diagonal with variance components $\nu_t$ on the diagonal and 0 off diagonal. This is what I assume to be the true form of D for now.  Remember that for every $\nu_t$, we can construct a matrix $E_t$ (with dimensions the same as matrix $D$) that has 1s on the diagonal elements corresponding to the elements of D that contain $\nu_t$ and 0s everywhere else.

We can partition the random effects according to their variance components: $U=(U_1',...,U_T')'$.  Let $D_t$ be the variance matrix for $U_t$ . $D_t$ has $q_t$ rows (and also columns). Thus $D$ is diagonal:
\begin{align}
D = \begin{bmatrix} D_1 & \cdots & 0 \\ \vdots & \ddots & \vdots \\ 0 & \cdots & D_T \end{bmatrix}
\end{align}

@@

The first and second derivatives of each summand with respect to its associated $\nu_t$ are:
\begin{align}
\dfrac{\partial}{\partial \nu_t} \log f_\theta(u_t) = - \dfrac{q_t}{2 \nu_t} + \dfrac{1}{2 \nu_t^2}(U_t-\mu_t)'(U_t-\mu_t)
\end{align}
and 
\begin{align}
\dfrac{\partial^2}{\partial \nu_t^2} \log f_\theta(u_t) = \dfrac{q_t}{2 \nu_t^2}- \dfrac{1}{\nu_t^3} (U_t-\mu_t)'(U_t-\mu_t).
\end{align}
Any other derivative is equal to 0. That is, for all $t_1 \neq t_2$,
\begin{align}
\dfrac{\partial}{\partial \nu_{t_1}} \log f_\theta(u_{t_2}) = 0.
\end{align}
Also,
\begin{align}
\dfrac{\partial}{\partial \beta} \log f_\theta(u_{t_2}) = 0.
\end{align}
Thus, if $\nu = (\nu_1,...,\nu_T)$, the gradient of the random effects distribution is the following vector of length $T$:
\begin{align}
\dfrac{\partial}{\partial \nu}  \log f_\theta(u) = \begin{bmatrix} - \dfrac{q_1}{2 \nu_1} + \dfrac{1}{2 \nu_1^2} (U_1-\mu_1) ' (U_1-\mu_1) & ... & - \dfrac{q_T}{2 \nu_T} + \dfrac{1}{2 \nu_T^2} (U_T-\mu_T) '(U_T-\mu_T)   \end{bmatrix} 
\end{align}
To calculate this vector, I will create a function to take $q_t$, $U_t$, and $\nu_t$ and  output $\dfrac{-q_t}{2\nu_t} +\dfrac{1}{2 \nu_t^2} U_t'U_t$. I can then do a loop for t in 1 through T to calculate each entry in the vector.

The Hessian matrix is the following diagonal matrix:
\begin{align}
\dfrac{\partial^2}{\partial \nu^2} \log f_\theta(u) = \begin{bmatrix} \dfrac{q_1}{2 \nu_1^2}- \dfrac{1}{\nu_1^3} U_1'U_1 & \cdots & 0 \\ \vdots & \ddots & \vdots \\ 0 & \cdots & \dfrac{q_T}{2 \nu_T^2}- \dfrac{1}{\nu_T^3 U_T'U_T} \end{bmatrix}
\end{align}
I will create another function to take $q_t$, $U_t$, $\mu_t$ and $\nu_t$ and spit out $\dfrac{q_t}{2 \nu_t^2} - \dfrac{1}{\nu_t^3}  (U_t-\mu_t)' (U_t-\mu_t)$. Again, during the above loop of length T, I will calculate the entries for the diagonal of this Hessian matrix.

The inputs of the function are the vector $\nu$ of variance components, U (the vector of random effects), and the list \texttt{z} (from \texttt{mod.mcml}).  The list \texttt{z} has $T$ matrices, each with the number of columns equal to $q_t$. We need $q_t,t=1,\ldots,T$ to calculate the log density and its derivatives.

@@

Rewritten in C, this function will take as pointers: the double array \texttt{nu} that contains the variance components, the int \texttt{T} to specify the length of \texttt{nu}, the double array \texttt{mu} that contains the means of the random effects, the int array \texttt{nrandom} of length \texttt{T} that contains the number random effects from that variance component, the double array \texttt{Uvec} that contains one vector of generated random effects,  the int array \texttt{meow} that specifies how to split \texttt{U} up based on the variance components, the double array \texttt{drgradient} that will contain the resulting gradient, and the double array \texttt{drhessian} that will contain the resulting hessian.

The result of invoking the C function will be the calculation of the gradient and hessian for the distribution of the random effects. The value will be evaluated by the C function in \ref{sec:Dgeneral}.

%%distRandGeneral
\subsubsection{$\Sigma$ in general (\texttt{distRandGeneral})}\label{sec:Dgeneral}
For now, this is only used for $\tilde{f}(u)$. (When $\Sigma$ is general, calculating the derivatives of $\log f_\theta(u)$ is not easy.) Recall the general form for a normal distribution with mean $\mu$ and variance $\Sigma$ is:
\begin{align}
\log f(u \,| \,\mu, \Sigma) = (1/2) \log |\Sigma^{-1}| - (1/2) (U-\mu)' \Sigma^{-1} (U-\mu)
\end{align}
 The only part of this to discuss is $|\Sigma^{-1}|$. We can use eigendecomposition to make $\Sigma^{-1}=O \Lambda O'$ where $O$ is orthogonal and $\Lambda$ is the diagonal matrix with eigenvalues. Since orthogonal matrices have determinant $\pm 1$, then $|O||O'|=1$. Thus  
\begin{align}
|\Sigma^{-1}|&=|O'| \; |\Lambda| \; |O| \\
&= |O'| \; |O| \; |\Lambda| \\
&=|\Lambda|,
\end{align}
which is just the product of the eigenvalues. The log of the determinant is calculated beforehand to save time, since this function is called $3m$ times each optimization iteration, where $m$ is again the Monte Carlo sample size.

@@

%\subsection{$\tilde{f}(u_k)$}\label{sec:ftwiddle}
%I need to evaluate $\log \tilde{f}(u_k)$ in the objective function of trust. I can make a function using the general form for a normal distribution's density as described in the previous subsection, then use that function for the three distributions that I sampled from. Since I generated the random effects from the three distributions in proportions $p_1,p_2,p_3$, I can write 
%\begin{align}
%\log \tilde{f}(u_k) = p_1 \log f(u_k \, | \, 0, D^*)+p_2 \log f(u_k \, | \, u^*, D^*)+p_3 \log f(u_k \, | \, u^*, (Z'  c''(X \beta^*+Zu^*) Z +(D^*)^{-1}   )^{-1})
%\end{align}
%
%\subsection{Distribution of random effects is multivariate t}\label{sec:tee}
%If we want, we can draw random effects from a t distribution with location parameter $u^*$ from PQL and scale matrix $D^*$ (using the variance components estimated by PQL). These can be drawn using the package mvtnorm. This package can also give the density evaluated at certain points. I'm not sure if I'll use this or just stick to normals.

@@

\section{Objective function: approximated log likelihood}\label{sec:objfun}
This will be hidden from the user. It will be called by the model fitting function.

In order to maximize the approximated log likelihood using trust, I will need an objective function that returns the value, the gradient vector, and the Hessian matrix of the log likelihood.  This objective function will get $\log f_\theta (u_k)$, $c(\eta_i)$, each of their gradient vectors, and each of their Hessian matrices to plug into these expressions.    Denote\\
\begin{align}
b_k &=  \log f_\theta (u_k,y)- \log \tilde{f} (u_k) \\
&= \log f_\theta (u_k) + \log f_\theta (y|u_k) - \log \tilde{f} (u_k) \label{eq:MCLA}
\end{align}

For computational stability, we'll set $a=max(b_k)$. Then the value is expressed as:\\
\begin{align}
\log f_{\theta,m} = a+ \log \left[ \dfrac{1}{m}  \sum_{k=1}^m e^{b_k-a} \right]
\end{align}

Define the weights as:\\
\begin{align}
v_\theta(u_k,y) = \dfrac{e^{b_k-a}}{ \sum_{k=1}^m e^{b_k-a}} 
\end{align}

The gradient vector is: \\
\begin{align}
G &= \sum_{k=1}^m \nabla \log f_\theta (u_k,y) v_\theta(u_k,y)\\ \label{eq:gradient}
&= \sum_{k=1}^m \nabla \left[  \log f_\theta(y|u_k) + \log f_\theta (u_k)  \right] v_\theta(u_k,y)\\
&= \sum_{k=1}^m  \left[ \nabla \log f_\theta(y|u_k) + \nabla\log f_\theta (u_k)  \right] v_\theta(u_k,y)\\
&= \sum_{k=1}^m  \left[ \dfrac{\partial}{\partial \beta} \log f_\theta(y|u_k) \; \; \; \;  \dfrac{\partial}{\partial \nu} \log f_\theta(y|u_k) \right] v_\theta(u_k,y)+ \left[ \dfrac{\partial}{\partial \beta} \log f_\theta(u_k) \; \; \; \; \dfrac{\partial}{\partial \nu} \log f_\theta(u_k) \right] v_\theta(u_k,y)\\
&= \sum_{k=1}^m  \left[ \dfrac{\partial}{\partial \beta} \log f_\theta(y|u_k) \; \; \; \; 0 \right] v_\theta(u_k,y)+ \left[ 0 \; \; \; \; \dfrac{\partial}{\partial \nu} \log f_\theta(u_k) \right] v_\theta(u_k,y)\\
&= \sum_{k=1}^m  \left[ \dfrac{\partial}{\partial \beta} \log f_\theta(y|u_k) \; \; \; \; \dfrac{\partial}{\partial \nu} \log f_\theta(u_k) \right] v_\theta(u_k,y)
\end{align}

>>>>

Then the Monte Carlo log likelihood approximation is
\begin{align}
l_{m}(\theta) &=\log \dfrac{1}{m} \sum_{k=1}^mf_\theta(y|u_k)  \dfrac{ f_\theta(u_k)   }{\tilde{f}(u_k)}\\
&= \log \dfrac{1}{m} \sum_{k=1}^m  \dfrac{ f_\theta(y,u_k)   }{\tilde{f}(u_k)}. \label{eq:MCLAval}
\end{align}
When $\tilde{f}$ depends on $\theta$, the gradient vector of the MCLA with respect to $\theta$ is
\begin{align}
\nabla l_m(\theta)&= \dfrac{\sum_{k=1}^m    \left( \nabla \log f_\theta(u_k,y) - \nabla \log \tilde{f} (u_k)  \right) \dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)} }{\sum_{k=1}^m \left( \dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)} \right) }, \label{eq:MCLAgradient}
\end{align}
and the Hessian matrix of the MCLA is
\begin{align}
&\nabla^2 l_m(\theta)= \dfrac{   \sum_{k=1}^m \left[ \nabla^2 \log f_\theta(y,u_k)  -
   \nabla^2 \log \tilde{f}(u_k)   \right]  \dfrac{ f_\theta(y,u_k)}{\tilde{f}(u_k)}  }{\sum_{k=1}^m  \dfrac{ f_\theta(y,u_k)   }{\tilde{f}(u_k)}}\\
&+ \dfrac{   \sum_{k=1}^m \left[ \nabla \log f_\theta(y,u_k)  -
   \nabla \log \tilde{f}(u_k) - \nabla l_m(\theta)   \right] \left[ \nabla \log f_\theta(y,u_k)  -
   \nabla \log \tilde{f}(u_k) -\nabla l_m(\theta)  \right]'  \dfrac{ f_\theta(y,u_k)   }{\tilde{f}(u_k)}   }{\sum_{k=1}^m  \dfrac{ f_\theta(y,u_k)   }{\tilde{f}(u_k)}}. \label{eq:MCLAhessian}
\end{align}
Details for these equations can be found in section \ref{sec:calcs}. Calculation for the gradient and hessian when $\tilde{f}$ is independent of $\theta$ can be found in section \ref{sec:calcsindep}.

@@

\section{Random effect generation and calculations}
This section is focused on a vector of random effects $u$ with length $q$. We assume that $u \sim N(0,D)$ where $D$ depends on $\nu$. Section \ref{sec:getEk} details the  relationship between  $\nu$ and $D$ explicit. Section \ref{sec:genRand} describes the process of generating the random effects. Section \ref{sec:distRand} explains how to evaluate the distribution of the random effects and its first two derivatives.

@@

If $q_t$ is the number of random effects associated with variance component $\nu_t$, then $q_t$ is also the number of nonzero entries in $E_t$ and $q=\sum_{t=1}^T q_t$ is the total number of random effects in the model.  
The \texttt{model.matrix} part of \texttt{glmm} has made  \texttt{Z} into a list with $T$ design matrices (one for each distinct variance component).  In order to create $E_t$, we need to go through and count the  number of columns ($q_t$) for each design matrix $t$ in the list.
 Then $q=\sum_{t=1}^T q_t$ is the total number of random effects in the model, and $D$  has $q$ rows (and columns). Thus, each $E_t$  has $q$ rows and columns as well.  Then we know that $E_1$ has $q_1$ ones on the diagonal, followed by zeros to fill the rest of the diagonal.  $E_2$ has $q_1$ zeros, then $q_2$ ones, then zeros for the rest of the diagonal.

\subsection{Generating random effects (genRand)}\label{sec:genRand}
The model fitting function \texttt{glmm}  calls the function that generates the random effects and  stores them for future calculations. Let $s$ be a vector of length $q$ that represents the random effects on the standard normal scale. That is, $u= D^{1/2}s$. Let $\sigma$ be a vector of length $T$ with components $\sqrt{\nu_t}$.  The  PQL step of \texttt{glmm} has resulted in PQL estimates for $\beta$, $\sigma$ and  $s$ denoted by $\beta^*$, $s^*$ and $\sigma^*$. Then let
\begin{align}
A^*=\sum_{t=1}^T E_t \sigma^*_t
\end{align}\
and
\begin{align}
 D^* =A^*A^*
\end{align}
be matrices based on PQL estimates. Also, we can ``unstandardize'' our PQL-based random effects:
\begin{align}
u^*&=A^*s^*.
\end{align}

Let $p_1,p_2,p_3$ be proportions such that  $p_1+p_2+p_3=1$. Let $D$ be determined by the current iteration of \texttt{trust}. The importance sampling distribution $\tilde{f}$ is the following mixture distribution:
\begin{align}
 \tilde{f}(u) = p_1  N(u \, | \, 0, D)+p_2  N(u \, | \, u^*, D^*)+p_3  N(u \, | \, u^*, (Z'  c''(X \beta^*+Zu^*) Z +(D^*)^{-1}   )^{-1}) \label{eq:ftwiddle}
\end{align}

@@

The random effects are generated by shifting and scaling a sample from a  standard normal distribution.  Suppose we want to generate a sample from a general normal distribution $N(\mu,\Sigma)$.  Let $u_k$ be a vector with length $q$ and let it denote one vector of generated random effects.  For each $k$, $k=1,\ldots,m$, draw a sample of size $q$ from the standard normal distribution. Call this sample $\breve{u}_k$. Then shift and scale it:
\begin{align}
u_k = \breve{u}_k \Sigma^{1/2} + \mu
\end{align}
so that it is a sample from $N(\mu,\Sigma)$.

This method requires $\Sigma^{1/2}$. One way to find $\Sigma^{1/2}$ is through eigendecomposition. Eigendecompositions take a little bit more time than Cholesky decompositions but are more stable. We can use eigendecomposition for $\Sigma$ to get orthogonal matrix $O$ (containing the eigenvectors) and diagonal matrix $\Lambda$ (with diagonal entries $\lambda$ being the eigenvalues of $\Sigma$).  Then
\begin{align}
\Sigma^{1/2}= O \Lambda^{1/2} O'.
\end{align}
Finding $\Lambda^{1/2}$ is as easy as taking the root of the diagonal entries.  We know that $\Sigma^{1/2}$ is correct because
\begin{align}
\Sigma^{1/2} \Sigma^{1/2} &= O \Lambda^{1/2} O'O \Lambda^{1/2} O'\\
&= O \Lambda^{1/2}\Lambda^{1/2} O'\\
&=O \Lambda O' \\
&= \Sigma.
\end{align}

@@

Consider $N(\mu, \Sigma)$.  This is written with $\mu$ not necessarily $0$ so that the function can be used for $\tilde{f}(u)$ as well.  We can  write the distribution  as:
\begin{align}
\log f (u| \mu, \Sigma) = (-1/2) \log |\Sigma| - (1/2) (u-\mu)' \Sigma^{-1} (u-\mu)
\end{align}

Since the first version of the package assumes $D$ is diagonal, formulas involving $D$ are found in section \ref{sec:Ddiag}.  
Since random effects are also generated from a normal with nondiagonal variance matrix, we also need to evaluate $f(u|\mu,\Sigma)$ for general covariance matrix $\Sigma$. Information on this calculation for general covariance matrix $\Sigma$ are in \ref{sec:Dgeneral}).

\subsubsection{$\Sigma=D$ Diagonal}\label{sec:Ddiag}
This section provides the equations for $\log f_\theta(u)$ and  $\log \tilde{f}(u)$ and their first two derivatives.

 Recall that for every $\nu_t$, we can construct a matrix $E_t$ (with dimensions the same as matrix $D$) that has 1s on the diagonal elements corresponding to the elements of D that contain $\nu_t$ and 0s  elsewhere.

We can partition the random effects according to their variance components: $u=(U_1',...,U_T')'$.  Let $D_t$ be the variance matrix for $U_t$. $D_t$ has $q_t$ rows (and also columns). Thus $D$ can be expressed as:
\begin{align}
D = \begin{bmatrix} D_1 & \cdots & 0 \\ \vdots & \ddots & \vdots \\ 0 & \cdots & D_T \end{bmatrix}
\end{align}

@@

The first and second derivatives of each summand with respect to its associated $\nu_t$ are:
\begin{align}
\dfrac{\partial}{\partial \nu_t} \log f_\theta(u_t) = - \dfrac{q_t}{2 \nu_t} + \dfrac{1}{2 \nu_t^2}(U_t-\mu_t)'(U_t-\mu_t)
\end{align}
and 
\begin{align}
\dfrac{\partial^2}{\partial \nu_t^2} \log f_\theta(u_t) = \dfrac{q_t}{2 \nu_t^2}- \dfrac{1}{\nu_t^3} (U_t-\mu_t)'(U_t-\mu_t).
\end{align}
Any other derivative is equal to 0. That is, for all $t_1 \neq t_2$,
\begin{align}
\dfrac{\partial}{\partial \nu_{t_1}} \log f_\theta(u_{t_2}) = 0.
\end{align}
Also,
\begin{align}
\dfrac{\partial}{\partial \beta} \log f_\theta(u_{t_2}) = 0.
\end{align}
Thus, if $\nu = (\nu_1,...,\nu_T)$, the gradient of the random effects distribution is the following vector of length $T$:
\begin{align}
\dfrac{\partial}{\partial \nu}  \log f_\theta(u) = \begin{bmatrix} - \dfrac{q_1}{2 \nu_1} + \dfrac{1}{2 \nu_1^2} (U_1-\mu_1) ' (U_1-\mu_1) & ... & - \dfrac{q_T}{2 \nu_T} + \dfrac{1}{2 \nu_T^2} (U_T-\mu_T) '(U_T-\mu_T)   \end{bmatrix} 
\end{align}

The Hessian matrix is the following diagonal matrix:
\begin{align}
\dfrac{\partial^2}{\partial \nu^2} \log f_\theta(u) = \begin{bmatrix} \dfrac{q_1}{2 \nu_1^2}- \dfrac{1}{\nu_1^3} U_1'U_1 & \cdots & 0 \\ \vdots & \ddots & \vdots \\ 0 & \cdots & \dfrac{q_T}{2 \nu_T^2}- \dfrac{1}{\nu_T^3 U_T'U_T} \end{bmatrix}
\end{align}
%I will create another function to take $q_t$, $U_t$, $\mu_t$ and $\nu_t$ and spit out $\dfrac{q_t}{2 \nu_t^2} - \dfrac{1}{\nu_t^3}  (U_t-\mu_t)' (U_t-\mu_t)$. Again, during the above loop of length T, I will calculate the entries for the diagonal of this Hessian matrix.

To calculate the value, gradient, and Hessian, we need to provide  $\nu$, $u$, and the list \texttt{z} (from \texttt{mod.mcml}).  The list \texttt{z} has $T$ matrices, each with the number of columns equal to $q_t$. We need $q_t,t=1,\ldots,T$ to calculate the log density and its derivatives.

@@

Rewritten in C, this function takes the following as pointers: the double array \texttt{nu} that contains the variance components, the int \texttt{T} to specify the length of \texttt{nu}, the double array \texttt{mu} that contains the means of the random effects, the int array \texttt{nrandom} of length \texttt{T} that contains the number random effects from that variance component, the double array \texttt{Uvec} that contains one vector of generated random effects,  the int array \texttt{meow} that specifies how to split \texttt{u} up based on the variance components, the double array \texttt{drgradient} that  contains the resulting gradient, and the double array \texttt{drhessian} that  contains the resulting hessian.

The result of invoking the C function is  the calculation of the gradient and hessian for the distribution of the random effects. The value is  evaluated by the C function in \ref{sec:Dgeneral}.

%%distRandGeneral
\subsubsection{$\Sigma$ in general (\texttt{distRandGeneral})}\label{sec:Dgeneral}
 Recall the general form for a normal distribution with mean $\mu$ and variance $\Sigma$ is:
\begin{align}
\log f(u \,| \,\mu, \Sigma) = (1/2) \log |\Sigma^{-1}| - (1/2) (u-\mu)' \Sigma^{-1} (u-\mu)
\end{align}
 The only part of this to discuss is $|\Sigma^{-1}|$. We can use eigendecomposition to make $\Sigma^{-1}=O \Lambda O'$ where $O$ is orthogonal and $\Lambda$ is the diagonal matrix with eigenvalues. Since orthogonal matrices have determinant $\pm 1$, then $|O||O'|=1$. Thus  
\begin{align}
|\Sigma^{-1}|&=|O'| \; |\Lambda| \; |O| \\
&= |O'| \; |O| \; |\Lambda| \\
&=|\Lambda|,
\end{align}
which is just the product of the eigenvalues. The log of the determinant is calculated beforehand to save time, since this function is called $3m$ times each optimization iteration, where $m$ is again the Monte Carlo sample size.

@@

%\subsection{$\tilde{f}(u_k)$}
%I need to evaluate $\log \tilde{f}(u_k)$ in the objective function of trust. I can make a function using the general form for a normal distribution's density as described in the previous subsection, then use that function for the three distributions that I sampled from. Since I generated the random effects from the three distributions in proportions $p_1,p_2,p_3$, I can write 
%\begin{align}
%\log \tilde{f}(u_k) = p_1 \log f(u_k \, | \, 0, D^*)+p_2 \log f(u_k \, | \, u^*, D^*)+p_3 \log f(u_k \, | \, u^*, (Z'  c''(X \beta^*+Zu^*) Z +(D^*)^{-1}   )^{-1})
%\end{align}
%
%\subsection{Distribution of random effects is multivariate t}\label{sec:tee}
%If we want, we can draw random effects from a t distribution with location parameter $u^*$ from PQL and scale matrix $D^*$ (using the variance components estimated by PQL). These can be drawn using the package mvtnorm. This package can also give the density evaluated at certain points. I'm not sure if I'll use this or just stick to normals.

@@

\subsection{Importance sampling distribution $\tilde{f}(u_k)$}\label{sec:ftwiddle}
Recall that the importance sampling distribution, first stated explicitly in equation \ref{eq:ftwiddle}, is:
\begin{align}
 \tilde{f}(u) = p_1  N(u \, | \, 0, D)+p_2  N(u \, | \, u^*, D^*)+p_3  N(u \, | \, u^*, (Z'  c''(X \beta^*+Zu^*) Z +(D^*)^{-1}   )^{-1}) 
\end{align}
Then
\begin{align}
 \log \tilde{f}(u) = \log \left( p_1  N(u \, | \, 0, D)+p_2  N(u \, | \, u^*, D^*)+p_3  N(u \, | \, u^*, (Z'  c''(X \beta^*+Zu^*) Z +(D^*)^{-1}   )^{-1}) \right)
\end{align}
Because only the first component of the mixture distribution contains any parameters,
\begin{align}
\dfrac{\partial}{\partial \nu} \log \tilde{f}(u) = \dfrac{p_1  \dfrac{\partial}{\partial \nu} N(u \, | \, 0, D)}{ p_1  N(u \, | \, 0, D)+p_2  N(u \, | \, u^*, D^*)+p_3  N(u \, | \, u^*, (Z'  c''(X \beta^*+Zu^*) Z +(D^*)^{-1}   )^{-1})}.
\end{align}
and 
\begin{align}
\nabla  \log \tilde{f}(u) = \left(0   \; \; \; \; \; \; \dfrac{\partial}{\partial \nu} \log \tilde{f}(u) \right)
\end{align}
These are used in equations \ref{eq:MCLAgr} and \ref{eq:MCLAh}. The second derivatives, which are needed in equation \ref{eq:MCLAh}, are
\begin{align}
\dfrac{\partial^2}{\partial \nu^2} \log \tilde{f}(u) &= \dfrac{p_1  \dfrac{\partial^2}{\partial \nu^2} N(u \, | \, 0, D)}{ p_1  N(u \, | \, 0, D)+p_2  N(u \, | \, u^*, D^*)+p_3  N(u \, | \, u^*, (Z'  c''(X \beta^*+Zu^*) Z +(D^*)^{-1}   )^{-1})} \\ \nonumber
&- \dfrac{p_1^2 \left( \dfrac{\partial}{\partial \nu} N(u \, | \, 0, D)\right) \left( \dfrac{\partial}{\partial \nu} N(u \, | \, 0, D)\right)'}{\left[ p_1  N(u \, | \, 0, D)+p_2  N(u \, | \, u^*, D^*)+p_3  N(u \, | \, u^*, (Z'  c''(X \beta^*+Zu^*) Z +(D^*)^{-1}   )^{-1})\right]^2}.
\end{align}
and
\begin{align}
\nabla^2 \log \tilde{f}(u_k) = \begin{bmatrix} 0 & 0 \\ 0  & \dfrac{\partial^2}{ \partial \nu^2} \log \tilde{f} (u_k)  \end{bmatrix} 
\end{align}

\section{Objective function: approximated log likelihood}\label{sec:objfun}
The objective function is optimized by \texttt{trust} within the the model fitting function.  In order to evaluate and maximize the approximated log likelihood, I need an objective function that returns the value, the gradient vector, and the Hessian matrix of the approximated log likelihood.  Recall equations \ref{eq:MCLAval}, \ref{eq:MCLAgradient}, \ref{eq:MCLAhessian}   for calculating these quantities.  This objective function needs $\log f_\theta (u_k)$, $c(\eta_i)$, $\tilde{f}(u_k)$, each of their gradient vectors, and each of their Hessian matrices to plug into these expressions.

Denote
\begin{align}
b_k &=  \log f_\theta (u_k,y)- \log \tilde{f} (u_k) \\
&= \log f_\theta (u_k) + \log f_\theta (y|u_k) - \log \tilde{f} (u_k) \label{eq:MCLA}
\end{align}

For computational stability, we'll set $a=max(b_k)$. Then the value of the MCLA is expressed as:\\
\begin{align}
\l_m (\theta) = a+ \log \left[ \dfrac{1}{m}  \sum_{k=1}^m e^{b_k-a} \right]
\end{align}

Define the weights as:
\begin{align}
v_\theta(u_k,y) = \dfrac{e^{b_k-a}}{ \sum_{k=1}^m e^{b_k-a}} 
\end{align}
